ODGI: understanding pangenome graphs

### 2022-03-14 manuscript/main.tex
defusing the bomb

\REVIEWED{Understanding pangenome graphs is not trivial. In the following, we highlight how ODGI's tools can assist in exploring the biological features of any pangenome graph.}

\REVIEWED{Pangenome graphs built from hundreds of haplotype-resolved \textit{de novo} genome assemblies are very large, but it is often necessary to work with only a small portion of the genomes represented,
such as a specific \textit{locus} (Fig.~\ref{fig:viz_extract_untangle-MHC}) or a smaller region (Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage,fig:viz_extract_untangle-C4-bandage,fig:viz_extract_untangle-C4-bandage-zoom}), or even a single gene (Fig.~\ref{fig:metrics}).
This simplifies the downstream analyses and reduces the resources to work only with the extracted graphs. More on graph extraction and edit operations can be found at $\S$\ref{sec:supp_edit}.}
%Subsetting operations allow us to zoom in on particular genes or regions of interest for more precise examination.

@@

\REVIEWED{Pangenome graph visualization is one of the first steps to gain insight into the mutual relationship between the sequences in the graph and their variation, understanding the higher-level biology of a graph.}
We pursue a novel approach to visualization with \textit{odgi draw} and \textit{odgi viz}, two tools that provide scalable ways of generating \REVIEWED{static} pictures of the high-level structure of even large pangenome graphs (Fig.~\ref{fig:odgi_viz}).

\REVIEWED{Using \textit{odgi extract}, we extracted the the major histocompatibility complex (MHC) locus from a 90 haplotypes human pangenome chromosome 6 graph form the HPRC. Specifically the graph contains the human references GRCH38, CHM13, and the contigs of 44 individuals that encode all possible variation including telomeres and centromeres. The MHC genes are involved in antigen presentation, inflammation regulation, the complement system, and the innate and adaptive immune responses~\citep{Shiina2009}.
MHC genes are highly polymorphic, i.e., there are multiple different alleles across individuals in a population.
Such variability becomes evident when we apply \textit{odgi draw} to visualize the graph layout of a human MHC pangenome graph (Fig.~\ref{fig:viz_extract_untangle-MHC}).
The visualization displays the graph topology in 2 dimensions (2D) and how the paths traverse the graph, with the regions of variations that appear as bubbles in the layout.
The layout was built by \textit{odgi layout} (see \S\ref{sec:sort}).} Of note, a 2D rendering can be costly to compute, but we provide an implementation that scales linearly with pangenome sequence size, allowing us to apply it to large pangenome graphs.
%\REVIEWED{For example, \textit{odgi draw} took 6 minutes and 21 seconds in total to produce the 2D layout of the MHC locus (Fig.~\ref{fig:odgi_layout}).}

@@

\REVIEWED{To obtain a more precise overview of the \textit{locus} in Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage},
we applied \textit{odgi untangle} with GRCH38 as a reference. \textit{odgi untangle} segments paths into linear segments by breaking these segments where the paths loop back on themselves.
In this way, we obtain information on the position and copy number status of the sequences in the collapsed \textit{locus} (Fig.~\ref{fig:viz_extract_untangle-C4-untangle}). The shape of the line indicates if the copy number is in forward or in reverse orientation compared to GRCH38. \textit{odgi untangle} demonstrates that it is able to work with any sets of reference sequences converting the graph to lift-over maps compatible with standard software for projecting annotations and alignments from one genome to another.  An explanation of the untangling process is given in $\S$\ref{sec:supp_navigation}.}
%This lets us decompose both homology and paralogy relationships by also untangling looping motifs in the graph, and obtaining linearized pairs of regions in BEDPE or PAF format.
%As an example, by untangling the graph we can study the variation that lies in regions collapsed due to ambiguous alignments over sequence repeats (as shown in Fig.~\ref{fig:viz_extract_untangle-C4-bandage}).

@@

\REVIEWED{Currently, bacterial pangenomes are best handled by specialized tools like PPanGGolin~\citep{Gautreau2020}, PanGraph~\citep{Noll2022} or PanX~\citep{Wei2017}. The latter one doesn't build a graphical representation of  a pangenome. But, it already has a very developed eco-system, which allows a detailed analysis of bacterial pangenomes using an interactive GUI. ODGI does not come with such sophisticated tooling, but it can serve as a strong basis to explore viral and bacterial pangenomes on a very large scale.
Other pangenome analysis platforms, like PanTools~\citep{Sheikhizadeh_2016} provide acces to pangenome analyses at the scales we demonstrate with ODGI, but use specialized de Bruijn graph models to achieve this---in contrast ODGI supports the highly generic variation graph model, which has greater representational power than de Bruijn graphs.}

>>>>

\REVIEWED{%Understanding pangenome graphs is not trivial.
Here, we apply our methods to a series of analyses, highlighting how ODGI can assist in exploring the biological features of pangenome graphs.
We follow typical analyses that we have found critical to interpreting whole genome alignments represented in the variation graph model.}

\REVIEWED{To simplify our exposition, we will extract small graph regions that are easy to interpret and describe.
Pangenome graphs built from hundreds of haplotype-resolved \textit{de novo} genome assemblies are very large, but it is often necessary to work with only a small portion of the genomes represented,
such as a specific locus (Fig.~\ref{fig:viz_extract_untangle-MHC}) or a smaller region (Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage,fig:viz_extract_untangle-C4-bandage,fig:viz_extract_untangle-C4-bandage-zoom}), or even a single gene (Fig.~\ref{fig:metrics}).
This simplifies the downstream analyses and reduces the resources to work only with the extracted graphs. More on graph extraction and edit operations can be found at $\S$\ref{sec:supp_edit}.}
%Subsetting operations allow us to zoom in on particular genes or regions of interest for more precise examination.

@@

\REVIEWED{Visualization methods help us quickly gain insight into otherwise opaque biological data.
We find visualization essential to understanding pangenome variation graphs.}
We pursue a novel approach to visualization with \textit{odgi draw} and \textit{odgi viz}, two tools that provide scalable ways of generating \REVIEWED{static} pictures of the high-level structure of even large pangenome graphs (Fig.~\ref{fig:odgi_viz}).

\REVIEWED{Using \textit{odgi extract}, we extracted the the major histocompatibility complex (MHC) locus from a 90 haplotypes human pangenome chromosome 6 graph form the HPRC. Specifically the graph contains the human references GRCH38, CHM13, and the contigs of 44 individuals that encode all possible variation including telomeres and centromeres. The MHC genes are involved in antigen presentation, inflammation regulation, the complement system, and the innate and adaptive immune responses~\citep{Shiina2009}.
MHC genes are highly polymorphic, i.e., there are multiple different alleles across individuals in a population.
Such variability becomes evident when we apply \textit{odgi draw} to visualize the graph layout of a human MHC pangenome graph (Fig.~\ref{fig:viz_extract_untangle-MHC}), using \textit{odgi layout} to generate the 2D projection (see \S\ref{sec:sort}).
The visualization displays the graph topology in 2 dimensions (2D), with structural variation that appears as bubbles in the layout.}
A 2D rendering can be costly to compute, but we provide an implementation that scales linearly with pangenome sequence size, allowing us to apply it to large pangenome graphs.
%\REVIEWED{For example, \textit{odgi draw} took 6 minutes and 21 seconds in total to produce the 2D layout of the MHC locus (Fig.~\ref{fig:odgi_layout}).}

@@

\REVIEWED{To obtain a more precise overview of the locus in Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage},
we applied \textit{odgi untangle} with GRCH38 as a reference. \textit{odgi untangle} segments paths into linear segments by breaking these segments where the paths loop back on themselves.
In this way, we obtain information on the position and copy number status of the sequences in the collapsed locus (Fig.~\ref{fig:viz_extract_untangle-C4-untangle}). The shape of the line indicates if the copy number is in forward or in reverse orientation compared to GRCH38. \textit{odgi untangle} demonstrates that it is able to work with any sets of reference sequences converting the graph to lift-over maps compatible with standard software for projecting annotations and alignments from one genome to another.  An explanation of the untangling process is given in $\S$\ref{sec:supp_navigation}.}
%This lets us decompose both homology and paralogy relationships by also untangling looping motifs in the graph, and obtaining linearized pairs of regions in BEDPE or PAF format.
%As an example, by untangling the graph we can study the variation that lies in regions collapsed due to ambiguous alignments over sequence repeats (as shown in Fig.~\ref{fig:viz_extract_untangle-C4-bandage}).

@@

\REVIEWED{Currently, bacterial pangenomes are best handled by specialized tools like PPanGGolin~\citep{Gautreau2020}, PanGraph~\citep{Noll2022} or PanX~\citep{Wei2017}. The latter one doesn't build a graphical representation of  a pangenome. But, it already has a very developed eco-system, which allows a detailed analysis of bacterial pangenomes using an interactive GUI. ODGI does not come with such sophisticated tooling, but it can serve as a strong basis to explore viral and bacterial pangenomes on a very large scale.
Other pangenome analysis platforms, like PanTools~\citep{Sheikhizadeh_2016} provide access to pangenome analyses at the scales we demonstrate with ODGI, but use specialized de Bruijn graph models to achieve this---in contrast ODGI supports the highly generic variation graph model, which has greater representational power than de Bruijn graphs.}
